---
title: "index"
author: "Rosie Hohnen"
date: "2024-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exploring threat overlap between woodland caribou and other 'species at risk' in Canada

Read in and clean data
```{r data_processing, echo=TRUE}
library(dplyr)

### read in dataset
threats <-read.csv('SAR_overlap_threats_noNA.csv')

### clean dataset
# Filter rows where percent_caribou > 0
overlap_caribou_threats <- threats %>%
  filter(Percent_caribou > 0)

# Replace spaces with underscores in a specific column (taxonomic_group)
overlap_caribou_threats <- overlap_caribou_threats %>%
  mutate(taxonomic_group = gsub(" ", "_", taxonomic_group))

# Keep only the most recent year_published for each common_name
overlap_caribou_threats_single <- overlap_caribou_threats %>%
  group_by(common_name) %>%
  filter(year_published == max(year_published)) %>%
  ungroup()

###delete caribou that aren't relevant and multiples of the same species that weren't eliminated with the process above
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(!(rowID %in% c("2024", "1505", "2047")))

### find rows with all zeros
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_total = rowSums(across(c("res_comm_development", "house_urban", "comm_industrial", "tourisim_rec", "agri_aqua", "crops", "wood_planta","livestock_farm", "aqua", "energy", "oil_gas", "mining", "renewable", "transport_service", "roads", "utility_service", "shipping_lane", "flight_path", "bio_resource_use", "hunting", "gathering", "logging", "fishing", "human_disturbance",
"recreation", "war", "work_activity", "natural_sys_change", "fire_fire_sup", "dams_water", "ecosysmod", "invasive_species",
"invasive_nonative", "problematic_native", "introduced_genetic", "problematic_species", "viral_disease", "disease_unknown",    
"pollution", "sewage_waste_water", "industrial_waste", "agricultural_waste", "garbage_soil_waste", "air_boure_pollute",
"ecess_energ7", "geological_event", "volcano", "earth_quake", "avalanche", "climate_change", "hab_shift", "drought", "temp_extreme","storm_flood"))))

###Delete all rows with zeros
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_total != "0") 

# Remove row 52 as sums to 0
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_number() != 52)

#### to make a consistent data set replace any 2 values with a 1
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(across(everything(), ~ ifelse(. == 2, 1, .)))

###Create subset using threat subcategories, and excluding overarching categories (that would be correlated with the sub categories)
subset_data <- overlap_caribou_threats_single %>% select(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40, 41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,68)

# Convert all columns to factors
subset_data <- subset_data %>% mutate(across(everything(), as.numeric))
```

## Run a MCA 

Examine patterns in how threats are shared between taxonomic groups, range areas and range overlap with caribou using an MCA.

```{r MCA, echo=TRUE}
### Run MCA to visualize differences based on taxonomic group
### try MCA
library("FactoMineR")
library("factoextra")

###Create subset for MCA. This subset has the sub groups but not the group categories
subset_data_MCA<- overlap_caribou_threats_single %>% select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,68))

# Remove rows and columns with no variance
subset_data_MCA <- subset_data_MCA[rowSums(is.na(subset_data_MCA)) != ncol(subset_data_MCA), ]
subset_data_MCA <- subset_data_MCA[, colSums(is.na(subset_data_MCA)) != nrow(subset_data_MCA)]

# Remove columns with very few non-zero values
threshold <- 3
subset_data_MCA <- subset_data_MCA %>%
  select(where(~ sum(.) >= threshold))

# Convert all columns to factors
subset_data_MCA <- data.frame(lapply(subset_data_MCA, as.factor))

#run MCA
MCA <-MCA(subset_data_MCA, ncp = 5, graph = FALSE)

eig.val <- get_eigenvalue(MCA)

# plot % variance explained
fviz_screeplot(MCA, addlabels = TRUE, ylim = c(0, 45))

var <- get_mca_var(MCA)
fviz_mca_var(MCA, col.var="black", shape.var = 15,
             repel = TRUE)

fviz_contrib(MCA, choice = "var", axes = 1, top = 15)# Contributions of rows to dimension 1
fviz_contrib(MCA, choice = "var", axes = 2, top = 15)# Contributions of rows to dimension 2
fviz_contrib(MCA, choice = "var", axes = 1:2, top = 15)# Total contribution to dimension 1 and 2

fviz_mca_var(MCA, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # avoid text overlapping (slow)
             ggtheme = theme_minimal()
)
### graph individuals (species in this case)

ind <- get_mca_ind(MCA)

```
# Plot species on the MCA in relation to taxonomic group, range overlap with acribou and range size

```{r MCA_plots2, echo=TRUE}
### visualize only individuals (species here)
fviz_mca_ind(MCA, col.ind = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, # Avoid text overlapping (slow if many points)
             ggtheme = theme_minimal())

### group by external variable 
# habillage = external grouping variable here taxonimic group
taxons <- as.factor(overlap_caribou_threats_single$taxonomic_group)
fviz_mca_ind(MCA, habillage = taxons, addEllipses = FALSE)

# Group by overlap with caribou
overlap <- as.numeric(overlap_caribou_threats_single$Percent_caribou)
fviz_mca_ind(
  MCA, 
  geom = "point", 
  col.ind = overlap,  # Use numeric vector for coloring
  gradient.cols = c("blue", "yellow", "red"),  # Continuous gradient
  addEllipses = FALSE
) +
  labs(color = "Overlap") # Adjust legend label

# Group by total range of caribou
range <- as.numeric(overlap_caribou_threats_single$Total_area_km)
fviz_mca_ind(
  MCA, 
  geom = "point", 
  col.ind = range, # Use Area for coloring
  gradient.cols = c("blue", "yellow", "red"), # Define a color gradient
  addEllipses = FALSE
) +
  labs(color = "Area") # Label for the color legend
```

## Calculate the Jaccard Index

Compute similarity between boreal caribou and other SAR in their IUCN threats using the Jaccard index.
```{r jaccard, echo=TRUE}

# Compute similarity between species and woodland caribou row using the Jaccard index

library(vegan)
dist <-vegdist(subset_data, method = "jaccard")
# Convert dist object to a matrix
dist_matrix <- as.matrix(dist)

# Convert matrix to a data frame
dist_df <- as.data.frame(dist_matrix)

# Extract row 42 from dist_df
row_17 <- dist_df[17, ]

# make similarity matrix
similarity_matrix <- 1 - dist_matrix

# Convert matrix to a data frame
sim_df <- as.data.frame(similarity_matrix)

# Extract row 42 from similarity matrix
row_17_sim <- sim_df[17, ]

# Add row_42 as a column to overlap_caribou_threats_single
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_17 = as.numeric(row_17))

# Add row_42_diss as a column to overlap_caribou_threats_single
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_17_sim = as.numeric(row_17_sim))

# Filter dataset for rows with similarity >0.48 to find ten most similar
jaccard_0.48 <- overlap_caribou_threats_single %>%
  filter(row_17_sim > 0.48)

# Delete threat columns
jaccard_0.48 <- jaccard_0.48 %>%
  select(-c(15,16,17,18,19,20,21,22,23,24,25, 26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,
            41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,
            67,68,69))

# find 10ish species most dissimilar
jaccard_0.21 <- overlap_caribou_threats_single %>%
  filter(row_17_sim < 0.21)

# Delete threat columns
jaccard_0.21 <- jaccard_0.21 %>%
  select(-c(15,16,17,18,19,20,21,22,23,24,25, 26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,
            41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,
            67,68,69))


```
## Calculate the number of columns in common between caribou and other species

How many columns overlap?
```{r common_columns, echo=TRUE}
####look at the number of column in common between caribou and all other species in the dataset
# First, extract the row corresponding to "caribou"
caribou_row <- overlap_caribou_threats_single %>% 
  filter(common_name == "caribou boreal population") %>%
  select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,
           41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,
           68))

# Create dataset to compare with
species_all_row <- overlap_caribou_threats_single %>%
  select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,
            41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,
            68))

# Create a new dataset comparing "caribou" to all other rows (and count 1's)
comparison <- species_all_row %>%
  rowwise() %>%
  mutate(matching_ones = sum(caribou_row == 1 & c_across(everything()) == 1, na.rm = TRUE)) %>%
  ungroup()

# Add common_columns column (!) to overlap_caribou_threats single
matching_ones <- comparison$matching_ones
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(matching_ones = as.numeric(matching_ones))

```

## Check for correlations prior to running models

Compute correlations between variables that could be included in the models. None appear to have correlation coefficients greater than 0.4.
```{r correlation, echo=TRUE}

library(corrplot)

model_subset <- overlap_caribou_threats_single %>% select(3,6,12,70,72,73,75,76,77)

numeric_vars <- model_subset %>%
  select(Percent_caribou, Total_area_km, sara_status, taxonomic_group, row_17, row_17_sim, matching_ones) %>%
  mutate(
  sara_status = as.numeric(as.factor(sara_status)),  # Convert factors to numeric
  taxonomic_group = as.numeric(as.factor(taxonomic_group))
  )

correlation_matrix <- cor(numeric_vars, use = "complete.obs")  # Use complete.obs to handle NAs
corrplot(correlation_matrix, method="circle")

```

## Model correlates of threat overlap with caribou

Run a glm that examine drivers of overlap in threats (Jaccard index) with caribou. These predictors include species range area (Total_area_km), percentage range overlap with caribou (Percent_caribou), taxonomic group (taxonomic_group_1 etc) and SARA status (sara_status1).
Quasibinomial model was chosen as the response variable (Jaccard index) goes from 0-1.

```{r model, echo=TRUE}

# Run some glms looking at descriptors of threat overlap with caribou
library(stats)
library(gtools)
library(MuMIn)

## Run some glms looking at descriptors of threat overlap with caribou
# Delete row 17 (caribou)
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_number() != 17)

# Make reference level for the variable taxonomic group
overlap_caribou_threats_single$taxonomic_group <- factor(overlap_caribou_threats_single$taxonomic_group, levels = c("Molluscs", "Arthropods", "Birds", "Lichens", "Amphibians","Mammals_(terrestrial)", "Mosses", "Reptiles", "Vascular_Plants"))

overlap_caribou_threats_single$taxonomic_group <- relevel(overlap_caribou_threats_single$taxonomic_group, ref = "Mosses")

b1 <- glm(round(cbind(row_17_sim, 1)*100)~ scale(Percent_caribou), data=overlap_caribou_threats_single, family=binomial)
b2 <- glm(round(cbind(row_17_sim, 1)*100)~ scale(Total_area_km), data=overlap_caribou_threats_single,family=binomial)
b3 <- glm(round(cbind(row_17_sim, 1)*100)~ sara_status, data=overlap_caribou_threats_single,family=binomial)
b4 <- glm(round(cbind(row_17_sim, 1)*100)~ taxonomic_group, data=overlap_caribou_threats_single,family=binomial)
b5 <- glm(round(cbind(row_17_sim, 1)*100)~ scale(Percent_caribou) + scale(Total_area_km) + sara_status + taxonomic_group, data=overlap_caribou_threats_single,family=binomial)
b6 <- glm(round(cbind(row_17_sim, 1)*100)~ 1, data=overlap_caribou_threats_single,family=binomial)

fmList3<-model.sel(b1=b1, b2=b2, b3=b3, b4=b4, b5=b5, b6=b6)
fmList3
summary(b5)

```

Evidence of greater threat overlap with caribou for species with greater range sizes, more overlap in their range with caribou, and for mammals and birds.
```{r check fit, echo=TRUE}

# Check fit
null_dev <- summary(b5)$null.deviance
resid_dev <- summary(b5)$deviance

prop_deviance_explained <- (null_dev - resid_dev) / null_dev
prop_deviance_explained

# Residuals vs. Fitted values
plot(fitted(b5), residuals(b5), main = "Residuals vs. Fitted")
abline(h = 0, col = "red")

# Normal Q-Q Plot of residuals
qqnorm(residuals(b5))
qqline(residuals(b5), col = "blue")

# Check fit
null_dev <- summary(b5)$null.deviance
resid_dev <- summary(b5)$deviance

prop_deviance_explained <- (null_dev - resid_dev) / null_dev
prop_deviance_explained

# Residuals vs. Fitted values
plot(fitted(b5), residuals(b5), main = "Residuals vs. Fitted")
abline(h = 0, col = "red")

# Normal Q-Q Plot of residuals
qqnorm(residuals(b5))
qqline(residuals(b5), col = "blue")

```
## Plot model averaged coefficient estimates of best fitting model

Currently b5
```{r coef, echo=TRUE}

# Extract coefficients and standard errors from the model
model_summary <- summary(b5)
coefficients <- as.data.frame(model_summary$coefficients)

# Create a new data frame for plotting
comp2 <- data.frame(
  Variable = rownames(coefficients),
  Estimate = coefficients[, "Estimate"],
  Std_Error = coefficients[, "Std. Error"],
  Lower = coefficients[, "Estimate"] - 1.96 * coefficients[, "Std. Error"],
  Upper = coefficients[, "Estimate"] + 1.96 * coefficients[, "Std. Error"])

# Rename categories
comp2 <- comp2 %>%
  mutate(Variable = recode(Variable, 
                         "scale(Percent_caribou)" = "Percentage range overlap with caribou",
                         "scale(Total_area_km)" = "Total range area",
                         "sara_statusSpecial Concern" = "Special concern",
                         "sara_statusThreatened" = "Threatened",
                         "taxonomic_groupArthropods" = "Arthropods",
                         "taxonomic_groupBirds" = "Birds",
                         "taxonomic_groupLichens" = "Lichens",
                         "taxonomic_groupAmphibians" = "Amphibians",
                         "taxonomic_groupMammals_(terrestrial)" = "Mammals (terrestrial)",
                         "taxonomic_groupReptiles" = "Reptiles",
                         "taxonomic_groupVascular_Plants" = "Vascular plants",
                         "taxonomic_groupMolluscs" = "Molluscs",
                          "(Intercept)"="Intercept"))


# Specify the desired order of categories
desired_order5 <- c("Total range area",
                    "Percentage range overlap with caribou",
                    "Threatened",
                    "Special concern",
                    "Mammals (terrestrial)",
                    "Amphibians",
                    "Reptiles",
                    "Molluscs",
                    "Birds",
                    "Vascular plants",
                    "Lichens",
                    "Arthropods",
                    "Intercept")
 
# Reorder the factor levels in the data frame
comp2$Variable <- factor(comp2$Variable, levels = desired_order5)

library(ggplot2)
p = ggplot(comp2,aes(x=Variable,y=Estimate)) +
  geom_hline(yintercept = 0, col = "darkgrey") +  # Horizontal line at y = 0
  geom_point() +
  geom_errorbar(aes(ymin=Lower,ymax=Upper)) +
  coord_flip() +
  labs(x= "Variable", y= "Coefficent estimate") +
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 30, b = 0, l = 0)),
    panel.background = element_rect(fill = 'white', colour = 'darkgrey'),
    panel.grid.major = element_line(colour = 'white'),
    text = element_text(size = 12),  # Increase overall font size
    axis.title = element_text(size = 14),  # Axis titles
    axis.text = element_text(size = 12))  # Axis text
print(p)

```
## Plot SAR taxonomic group and caribou range overlap

Make a graph that shows which taxonomic groups of SAR overlap most with caribou ranges
```{r overlap, echo=TRUE}
library(ggplot2)
overlap <- read.csv("SAR_cara_overlap.csv", header=TRUE,)

# Reshape data for grouped bar plot
library(tidyr)
overlap_long <- overlap %>%
  pivot_longer(cols = c(SAR, SAR_caribou_range, SAR_caribou_range_20),
               names_to = "Metric", 
               values_to = "Value")

# Specify the desired order of categories
desired_order4 <- c(
  "Vascular Plants",
  "Birds",
  "Arthropods",
  "Reptiles",
  "Amphibians",
  "Mammals",
  "Lichens",
  "Molluscs",
  "Mosses"
)

# Reorder the factor levels in the data frame
overlap_long$Group <- factor(overlap_long$Group, levels = desired_order4)

# Create grouped bar plot
ggplot(overlap_long, aes(x = Group, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Taxonomic group",
       y = "Number of species",
       fill = "Metric") +
  scale_fill_manual(values = c("darkseagreen2", "khaki1", "green4"),
                    labels = c("SAR", "SAR that overlap with caribou", "SAR that overlap with caribou >20%")) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 30, b = 0, l = 0)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = 'white', colour = 'darkgrey'),
    panel.grid.major = element_line(colour = 'white'),
    text = element_text(size = 16),  # Increase overall font size
    axis.title = element_text(size = 18),  # Axis titles
    axis.text = element_text(size = 16),  # Axis text
  )

```
## What are the most common threats to SAR in Canada?

How do these vary from the species within the caribou range, and those listed for caribou?
```{r common_threats, echo=TRUE}

### First make data set
# Filter rows where percent_caribou > 0

# Replace spaces with underscores in a specific column (taxonomic_group)
threats_all <- threats %>%
  mutate(taxonomic_group = gsub(" ", "_", taxonomic_group))

# Keep only the most recent year_published for each common_name
threats_all <- threats_all %>%
  group_by(common_name) %>%
  filter(year_published == max(year_published)) %>%
  ungroup()

###delete caribou that aren't relevant and multiples of the same species that weren't eliminated with the process above
threats_all <- threats_all %>%
  filter(!(rowID %in% c("1605", "631","1545", "732","615", "612", "639","663", "114",
                        "889", "662","221", "228", "909", "579", "697", "710", "320",
                        "543", "358", "923", "567")))

#### to make a consistent data set replace any 2 values with a 1
threats_all <- threats_all %>%
  mutate(across(everything(), ~ ifelse(. == 2, 1, .)))

library(tidyr)
subset_data_caribou_title <- overlap_caribou_threats_single %>% select(15,19,24,28,33,38,42,46,53,60,64,69)
subset_cat_title <- threats_all %>% select(15,19,24,28,33,38,42,46,53,60,64,69)

summed_threats_caribou <- subset_data_caribou_title %>% 
  summarise(across(everything(), sum)) %>%
  pivot_longer(everything(), names_to = "Threat", values_to = "Summed_Species_cara") 

summed_threats_title <- subset_cat_title %>% 
  summarise(across(everything(), sum)) %>%
  pivot_longer(everything(), names_to = "Threat", values_to = "Summed_Species_all")

Threats_both <- summed_threats_title %>%
  mutate(Summed_Species_cara = summed_threats_caribou$Summed_Species_cara)

## Calculate proportions of all species
Threats_both <- Threats_both %>%
  mutate(Proportion_all = (Summed_Species_all/612)*100)

## Calculate proportions of caribou overlapping species
Threats_both <- Threats_both %>%
  mutate(Proportion_cara = (Summed_Species_cara/91)*100)

Threats_both2 <- Threats_both %>% 
  pivot_longer(cols = starts_with("Proportion"), names_to = "type", values_to = "Proportion")

# Rename categories
Threats_both2 <- Threats_both2 %>%
  mutate(Threat = recode(Threat, 
                        "agri_aqua" = "Agriculture and aquaculture",
                        "bio_resource_use" = "Biological resource use",
                        "climate_change" = "Climate change",
                         "energy" = "Energy production and mining",
                         "geological_event" = "Geological events",
                         "human_disturbance" = "Human disturbance",
                         "invasive_species" = "Invasive and problematic species",
                         "natural_sys_change" = "Natural system modifications",
                         "other_impacts" = "Other options",
                         "pollution" = "Pollution",
                         "res_comm_development" = "Residential and commercial development",
                         "transport_service" = "Transport and service corridors"))

# Specify the desired order of categories
desired_order3 <- c(
  "Invasive and problematic species",
  "Natural system modifications",
  "Pollution",
  "Climate change",
  "Human disturbance",
  "Residential and commercial development",
  "Biological resource use",
  "Transport and service corridors",
  "Agriculture and aquaculture",
  "Energy production and mining",
  "Geological events",
  "Other options"
)
# Reorder the factor levels in the data frame
Threats_both2$Threat <- factor(Threats_both2$Threat, levels = desired_order3)

# Define custom colors
custom_colors <- c(
  'Proportion_all' = 'deepskyblue4',  # Dark teal color for 'wolverine'
  'Proportion_cara' = 'cadetblue2'  # Light blue color for 'woodland caribou'
)

# Create grouped bar plot for threats to all SAR
ggplot(Threats_both2, aes(x = Threat, y = Proportion, fill=type, width=0.45)) +
  geom_bar(stat = "identity", position='dodge') +
  labs(x = "Threat",
       y = "Proportion of species") +
  theme_minimal() +
  scale_fill_manual(values = custom_colors) +
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
    axis.text.x = element_text(angle = 65, hjust = 1, margin = margin(t = 0, r = 0, b = 20, l = 0)),
    panel.background = element_rect(fill = 'white', colour = 'darkgrey'),
    panel.grid.major = element_line(colour = 'white'),
    text = element_text(size = 16),  # Increase overall font size
    axis.title = element_text(size = 18),  # Axis titles
    axis.text = element_text(size = 16),  # Axis text
    legend.position = "none"
  )

```