---
title: "index"
author: "Rosie Hohnen"
date: "2024-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exploring threat overlap between woodland caribou and other 'species at risk' in Canada

Read in and clean data
```{r data_processing, echo=TRUE}
library(dplyr)

### read in dataset
threats <-read.csv('SAR_overlap_threats_noNA.csv')

### clean dataset
# Filter rows where percent_caribou > 0
overlap_caribou_threats <- threats %>%
  filter(Percent_caribou > 0)

# Replace spaces with underscores in a specific column (taxonomic_group)
overlap_caribou_threats <- overlap_caribou_threats %>%
  mutate(taxonomic_group = gsub(" ", "_", taxonomic_group))

# Keep only the most recent year_published for each common_name
overlap_caribou_threats_single <- overlap_caribou_threats %>%
  group_by(common_name) %>%
  filter(year_published == max(year_published)) %>%
  ungroup()

###delete caribou that aren't relevant and multiples of the same species that weren't eliminated with the process above
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(!(rowID %in% c("2024", "1505", "2047")))

### find rows with all zeros
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_total = rowSums(across(c("res_comm_development", "house_urban", "comm_industrial", "tourisim_rec", "agri_aqua", "crops", "wood_planta","livestock_farm", "aqua", "energy", "oil_gas", "mining", "renewable", "transport_service", "roads", "utility_service", "shipping_lane", "flight_path", "bio_resource_use", "hunting", "gathering", "logging", "fishing", "human_disturbance",
"recreation", "war", "work_activity", "natural_sys_change", "fire_fire_sup", "dams_water", "ecosysmod", "invasive_species",
"invasive_nonative", "problematic_native", "introduced_genetic", "problematic_species", "viral_disease", "disease_unknown",    
"pollution", "sewage_waste_water", "industrial_waste", "agricultural_waste", "garbage_soil_waste", "air_boure_pollute",
"ecess_energ7", "geological_event", "volcano", "earth_quake", "avalanche", "climate_change", "hab_shift", "drought", "temp_extreme","storm_flood"))))

###Delete all rows with zeros
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_total != "0") 

# Remove row 52 as sums to 0
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_number() != 52)

#### to make a consistent data set replace any 2 values with a 1
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(across(everything(), ~ ifelse(. == 2, 1, .)))

###Create subset using threat subcategories, and excluding overarching categories (that would be correlated with the sub categories)
subset_data <- overlap_caribou_threats_single %>% select(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40, 41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,68)

# Convert all columns to factors
subset_data <- subset_data %>% mutate(across(everything(), as.numeric))
```

## Run a MCA 

Examine patterns in how threats are shared between taxonomic groups, range areas and range overlap with caribou using an MCA.

```{r MCA, echo=TRUE}
### Run MCA to visualize differences based on taxonomic group
### try MCA
library("FactoMineR")
library("factoextra")

###Create subset for MCA. This subset has the sub groups but not the group categories
subset_data_MCA<- overlap_caribou_threats_single %>% select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,68))

# Remove rows and columns with no variance
subset_data_MCA <- subset_data_MCA[rowSums(is.na(subset_data_MCA)) != ncol(subset_data_MCA), ]
subset_data_MCA <- subset_data_MCA[, colSums(is.na(subset_data_MCA)) != nrow(subset_data_MCA)]

# Remove columns with very few non-zero values
threshold <- 3
subset_data_MCA <- subset_data_MCA %>%
  select(where(~ sum(.) >= threshold))

# Convert all columns to factors
subset_data_MCA <- data.frame(lapply(subset_data_MCA, as.factor))

#run MCA
MCA <-MCA(subset_data_MCA, ncp = 5, graph = FALSE)

eig.val <- get_eigenvalue(MCA)

# plot % variance explained
fviz_screeplot(MCA, addlabels = TRUE, ylim = c(0, 45))

var <- get_mca_var(MCA)
fviz_mca_var(MCA, col.var="black", shape.var = 15,
             repel = TRUE)

fviz_contrib(MCA, choice = "var", axes = 1, top = 15)# Contributions of rows to dimension 1
fviz_contrib(MCA, choice = "var", axes = 2, top = 15)# Contributions of rows to dimension 2
fviz_contrib(MCA, choice = "var", axes = 1:2, top = 15)# Total contribution to dimension 1 and 2

fviz_mca_var(MCA, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # avoid text overlapping (slow)
             ggtheme = theme_minimal()
)
### graph individuals (species in this case)

ind <- get_mca_ind(MCA)

```
# Plot species on the MCA in relation to taxonomic group, range overlap with acribou and range size

```{r MCA_plots2, echo=TRUE}
### visualize only individuals (species here)
fviz_mca_ind(MCA, col.ind = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, # Avoid text overlapping (slow if many points)
             ggtheme = theme_minimal())

### group by external variable 
# habillage = external grouping variable here taxonimic group
taxons <- as.factor(overlap_caribou_threats_single$taxonomic_group)
fviz_mca_ind(MCA, habillage = taxons, addEllipses = FALSE)

# Group by overlap with caribou
overlap <- as.numeric(overlap_caribou_threats_single$Percent_caribou)
fviz_mca_ind(
  MCA, 
  geom = "point", 
  col.ind = overlap,  # Use numeric vector for coloring
  gradient.cols = c("blue", "yellow", "red"),  # Continuous gradient
  addEllipses = FALSE
) +
  labs(color = "Overlap") # Adjust legend label

# Group by total range of caribou
range <- as.numeric(overlap_caribou_threats_single$Total_area_km)
fviz_mca_ind(
  MCA, 
  geom = "point", 
  col.ind = range, # Use Area for coloring
  gradient.cols = c("blue", "yellow", "red"), # Define a color gradient
  addEllipses = FALSE
) +
  labs(color = "Area") # Label for the color legend
```

## Calculate the Jaccard Index

Compute similarity between boreal caribou and other SAR in their IUCN threats using the Jaccard index.
```{r jaccard, echo=TRUE}

# Compute similarity between species and woodland caribou row using the Jaccard index

library(vegan)
dist <-vegdist(subset_data, method = "jaccard")
# Convert dist object to a matrix
dist_matrix <- as.matrix(dist)

# Convert matrix to a data frame
dist_df <- as.data.frame(dist_matrix)

# Extract row 42 from dist_df
row_17 <- dist_df[17, ]

# make similarity matrix
similarity_matrix <- 1 - dist_matrix

# Convert matrix to a data frame
sim_df <- as.data.frame(similarity_matrix)

# Extract row 42 from similarity matrix
row_17_sim <- sim_df[17, ]

# Add row_42 as a column to overlap_caribou_threats_single
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_17 = as.numeric(row_17))

# Add row_42_diss as a column to overlap_caribou_threats_single
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(row_17_sim = as.numeric(row_17_sim))

# find species with jaccard scores of >0.5

# Filter dataset for rows with similarity >0.5
jaccard_0.5 <- overlap_caribou_threats_single %>%
  filter(row_17_sim > 0.5)

# Delete threat columns
jaccard_0.5 <- jaccard_0.5 %>%
  select(-c(15,16,17,18,19,20,21,22,23,24,25, 26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,
            41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,
            67,68,69))

# look at species most dissimilar
jaccard_0.2 <- overlap_caribou_threats_single %>%
  filter(row_17_sim < 0.2)

# Delete threat columns
jaccard_0.2 <- jaccard_0.2 %>%
  select(-c(15,16,17,18,19,20,21,22,23,24,25, 26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,
            41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,
            67,68,69))

```
## Calculate the number of columns in common between caribou and other species

How many columns overlap?
```{r common_columns, echo=TRUE}
####look at the number of column in common between caribou and all other species in the dataset
# First, extract the row corresponding to "caribou"
caribou_row <- overlap_caribou_threats_single %>% 
  filter(common_name == "caribou boreal population") %>%
  select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,
           41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,
           68))

# Create dataset to compare with
species_all_row <- overlap_caribou_threats_single %>%
  select(c(16,17,18,20,21,22,23,25,26,27,29,30,31,32,34,35,36,37,39,40,
            41,43,44,45,47,48,49,50,51,52,54,55,56,57,58,59,61,62,63,65,66,67,
            68))

# Create a new dataset comparing "caribou" to all other rows (and count 1's)
comparison <- species_all_row %>%
  rowwise() %>%
  mutate(matching_ones = sum(caribou_row == 1 & c_across(everything()) == 1, na.rm = TRUE)) %>%
  ungroup()

# Add common_columns column (!) to overlap_caribou_threats single
matching_ones <- comparison$matching_ones
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  mutate(matching_ones = as.numeric(matching_ones))

```

## Check for correlations prior to running models

Compute correlations between variables that could be included in the models. None appear to have correlation coefficients greater than 0.4.
```{r correlation, echo=TRUE}

library(corrplot)

model_subset <- overlap_caribou_threats_single %>% select(3,6,12,70,72,73,75,76,77)

numeric_vars <- model_subset %>%
  select(Percent_caribou, Total_area_km, sara_status, taxonomic_group, row_17, row_17_sim, matching_ones) %>%
  mutate(
  sara_status = as.numeric(as.factor(sara_status)),  # Convert factors to numeric
  taxonomic_group = as.numeric(as.factor(taxonomic_group))
  )

correlation_matrix <- cor(numeric_vars, use = "complete.obs")  # Use complete.obs to handle NAs
corrplot(correlation_matrix, method="circle")

```

## Model correlates of threat overlap with caribou

Run a glm that examine drivers of overlap in threats (Jaccard index) with caribou. These predictors include species range area (Total_area_km), percentage range overlap with caribou (Percent_caribou), taxonomic group (taxonomic_group_1 etc) and SARA status (sara_status1).
Quasibinomial model was chosen as the response variable (Jaccard index) goes from 0-1.

```{r model, echo=TRUE}

# Run some glms looking at descriptors of threat overlap with caribou

library(stats)
# delete row 17 (carabou row)
overlap_caribou_threats_single <- overlap_caribou_threats_single %>%
  filter(row_number() != 17)

library(stats)
library(gtools)
m1 <- glm(row_17_sim~ scale(Percent_caribou) + scale(Total_area_km) + sara_status + taxonomic_group, 
          data=overlap_caribou_threats_single, family=quasibinomial(link = "logit"))
summary(m1)


m2 <- glm(cbind(matching_ones, 33-matching_ones)~ scale(Percent_caribou) + scale(Total_area_km) + sara_status + taxonomic_group, data=overlap_caribou_threats_single,
          family=binomial)
summary(m2)


```

Evidence of greater threat overlap with caribou for species with greater range sizes, more overlap in their range with caribou, and for mammals and birds.
```{r check fit, echo=TRUE}

# Check fit
null_dev <- summary(m1)$null.deviance
resid_dev <- summary(m1)$deviance

prop_deviance_explained <- (null_dev - resid_dev) / null_dev
prop_deviance_explained

# Residuals vs. Fitted values
plot(fitted(m1), residuals(m1), main = "Residuals vs. Fitted")
abline(h = 0, col = "red")

# Normal Q-Q Plot of residuals
qqnorm(residuals(m1))
qqline(residuals(m1), col = "blue")

# Check fit
null_dev <- summary(m2)$null.deviance
resid_dev <- summary(m2)$deviance

prop_deviance_explained <- (null_dev - resid_dev) / null_dev
prop_deviance_explained

# Residuals vs. Fitted values
plot(fitted(m2), residuals(m2), main = "Residuals vs. Fitted")
abline(h = 0, col = "red")

# Normal Q-Q Plot of residuals
qqnorm(residuals(m2))
qqline(residuals(m2), col = "blue")

```

## Plot SAR taxonomic group and caribou range overlap

Make a graph that shows which taxonomic groups of SAR overlap most with caribou ranges
```{r overlap, echo=TRUE}
library(ggplot2)
overlap <- read.csv("SAR_cara_overlap.csv", header=TRUE,)

# Reshape data for grouped bar plot
library(tidyr)
overlap_long <- overlap %>%
  pivot_longer(cols = c(SAR, SAR_caribou_range, SAR_caribou_range_20),
               names_to = "Metric", 
               values_to = "Value")

# Create grouped bar plot
ggplot(overlap_long, aes(x = Group, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Taxonomic group",
       y = "Number of species",
       fill = "Metric") +
  scale_fill_manual(values = c("darkseagreen2", "khaki1", "green4"),
                    labels = c("SAR", "SAR that overlap with caribou", "SAR that overlap with caribou >20%")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = 'white', colour = 'darkgrey'),
    panel.grid.major = element_line(colour = 'white'),
    text = element_text(size = 16),  # Increase overall font size
    axis.title = element_text(size = 18),  # Axis titles
    axis.text = element_text(size = 16),  # Axis text
  )

```


